% Chapter imported from the companion textbook
% (formal-learning-theory-book) with lean annotations added
% to cross-link each formalized statement to its kernel declaration.
% Prose is unchanged; annotations are additive.

% Chapter 7: Identification in the Limit
% ex_learning = Profile B (load-bearing), bc_learning = Profile C (revelation),
% mind_change_characterization = Profile B, online_pac_gold_separation = Profile D
%
% Texture: recursion-theoretic. Proofs use diagonalization, not concentration
% or combinatorial trees. Infinite horizon, no sample complexity, ordinal-valued
% mind-change bounds. This chapter should feel fundamentally different from
% Ch5 (PAC) and Ch6 (Online).

\chapter{Identification in the Limit}
\label{ch:gold}

In 1967, seventeen years before Valiant introduced PAC learning, E.\ Mark Gold posed a question that remains the oldest open research programme in computational learning theory: which classes of recursive functions can a machine identify from examples?

Gold's framework differs from PAC and online learning not merely in its definitions but in its \emph{proof technique}. PAC theory rests on concentration inequalities, the probabilistic convergence of empirical risk to true risk. Online learning rests on combinatorial game trees, the Littlestone dimension measures the depth of a binary tree that the adversary can force. Gold-style identification rests on \emph{diagonalization}, the recursion-theoretic technique of defeating every candidate learner by constructing an adversarial input that forces failure. The proof method shapes the entire chapter.

Three features distinguish this paradigm from everything that precedes it in this book:

\begin{enumerate}[leftmargin=*,itemsep=3pt]
\item \textbf{Infinite horizon.} The learner receives an infinite stream of data and must eventually converge. There is no sample complexity bound, no $\varepsilon$, no $\delta$. The question is whether the learner converges, not how fast.

\item \textbf{Ordinal-valued complexity.} When we ask ``how many times does the learner change its mind?'' the answer is not an integer but a \emph{transfinite ordinal}. This is the first appearance of ordinals in learning theory.

\item \textbf{A lattice of success criteria.} PAC learning has one definition (modulo agnostic, realizable, improper variants). Gold-style identification has a rich hierarchy: $\FIN < \Ex < \BC$, with anomalous, monotonic, and vacillatory learning branching off. The hierarchy itself is mathematical content.
\end{enumerate}

\begin{historical}
Gold's 1967 paper~\cite{Gold1967} was preceded by his 1965 paper on limiting recursion~\cite{Gold1965}, which introduced the idea that a computation could ``converge in the limit'' even if it never halts in the classical sense. The 1967 paper applied this idea to language learning. It was the first mathematical theory of learning from examples, predating Valiant's PAC model by 17 years, Vapnik and Chervonenkis's foundational work on uniform convergence by 4 years, and Littlestone's online model by 21 years. The diagonalization technique Gold used to prove his impossibility theorem later became standard in computational complexity theory, but Gold's use predates most of those applications.
\end{historical}

% ================================================================
% SECTION 1: Gold's Question
% ================================================================

\section{Gold's Question}
\label{sec:gold-question}

Fix a countable domain. In Gold's original setting, this is the set of all strings over a finite alphabet $\Sigma$, and the objects to be learned are formal languages $L \subseteq \Sigma^*$. We work in this setting when it clarifies the recursion-theoretic content, and in the general setting of concept classes $\mathcal{C} \subseteq 2^\N$ when it does not matter.

The learner receives data sequentially, one datum at a time, forever. Two data presentation modes are standard:

\begin{defn}[Text and Informant]
\label{def:text-informant}
Let $L \subseteq \Sigma^*$ be a language.
\begin{itemize}[leftmargin=*,itemsep=2pt]
\item A \emph{text} for $L$ is an infinite sequence $t = (t_0, t_1, t_2, \ldots)$ with $\{t_i : i \in \N\} = L$. Every element of $L$ appears at least once; no element of $\Sigma^* \setminus L$ ever appears. The text may contain repetitions and need not follow any fixed ordering.
\item An \emph{informant} for $L$ is an infinite sequence of labelled pairs $((s_0, b_0), (s_1, b_1), \ldots)$ where $\{s_i : i \in \N\} = \Sigma^*$ and $b_i = 1$ if and only if $s_i \in L$. Every string is eventually presented together with its membership status.
\end{itemize}
\end{defn}

Text presents only positive examples; informant presents both positive and negative examples. This distinction matters enormously: Gold showed that informant is strictly more powerful than text. We focus on text, which is the harder and more interesting case.

After seeing the first $n$ data points $(t_0, \ldots, t_{n-1})$, the learner outputs a hypothesis $h_n$. The hypothesis is an \emph{index}, a natural number naming a program that computes a language. The learner has no deadline, no accuracy target, and no probability distribution on targets. It simply sees more and more of $L$ and must eventually guess correctly.

% ================================================================
% SECTION 2: Ex-Learning and Gold's Impossibility Theorem
% ================================================================

\section{Ex-Learning and Gold's Impossibility Theorem}
\label{sec:ex-gold}

\begin{defn}[Explanatory Learning ($\Ex$)]
\label{def:ex-learning}
  \lean{EXLearnable}
  \leanok
A learner $M$ \emph{$\Ex$-identifies} a language $L$ from text if, for every text $t$ for $L$, there exists an index $e$ and a time $n_0$ such that $M(t_0, \ldots, t_n) = e$ for all $n \geq n_0$, and $W_e = L$ (where $W_e$ is the language computed by program $e$).

A class $\mathcal{L}$ of languages is \emph{$\Ex$-identifiable} if there exists a single learner $M$ that $\Ex$-identifies every $L \in \mathcal{L}$ from text.
\end{defn}

The definition requires \emph{syntactic convergence}: the learner must eventually output the same index forever. It is not enough to output a sequence of indices that all happen to compute the correct language, the index itself must stabilize. (Relaxing this requirement gives BC-learning; see \Cref{sec:bc-learning}.)

\begin{example}[$\Ex$-identification of finite languages]
\label{ex:finite-languages}
Let $\mathcal{L}_{\mathrm{fin}}$ be the class of all finite subsets of $\N$. Here is an $\Ex$-learner for $\mathcal{L}_{\mathrm{fin}}$: at time $n$, output an index for $\{t_0, \ldots, t_n\}$. After all elements of $L$ have appeared (which happens at some finite time $n_0$, since $L$ is finite), the set $\{t_0, \ldots, t_n\}$ equals $L$ for all $n \geq n_0$. The hypothesis stabilizes.
\end{example}

This simple algorithm illustrates a crucial asymmetry: the learner does not know \emph{when} it has converged. It has no way to announce ``I am done.'' It merely stabilizes, silently. This lack of a convergence signal is what makes Gold's impossibility theorem possible.

\begin{thm}[Gold's Impossibility Theorem {\cite{Gold1967}}]
\label{thm:gold}
  \lean{gold_theorem}
  \leanok
  \uses{def:ex-learning}
Let $\mathcal{L}$ be any class of languages that contains all finite languages and at least one infinite language. Then $\mathcal{L}$ is not $\Ex$-identifiable from text.
\end{thm}

This is the centerpiece of the chapter. The proof is a diagonalization argument: we defeat \emph{every} candidate learner by constructing a text that forces it to fail. The construction is adversarial, not probabilistic, there is no distribution, no $\varepsilon$-net, no covering number. The adversary simply watches the learner and manipulates the future of the stream to prevent convergence.

\begin{proof}
Let $M$ be any learner. We construct a text $t$ for some language $L \in \mathcal{L}$ such that $M$ fails to $\Ex$-identify $L$ on $t$.

We build $t$ in stages. Let $L_\infty \in \mathcal{L}$ be the infinite language that $\mathcal{L}$ contains by assumption. Enumerate $L_\infty = \{w_0, w_1, w_2, \ldots\}$.

\medskip
\noindent\textbf{Stage 0.} Present $w_0$ to $M$. The learner outputs some hypothesis $h_0 = M(w_0)$.

\medskip
\noindent\textbf{Stage $k \geq 1$.} At this point we have presented the finite sequence $(w_0, \ldots, w_{n_{k-1}})$ for some $n_{k-1}$, and $M$ has output hypothesis $h_{n_{k-1}}$. Let $F_k = \{w_0, \ldots, w_{n_{k-1}}\}$ be the set of strings presented so far.

Consider two cases:
\begin{itemize}[leftmargin=*]
\item \textbf{Case A:} $W_{h_{n_{k-1}}} = F_k$ (the current hypothesis computes exactly the finite set seen so far). Then present $w_{n_{k-1}+1}$ (the next element of $L_\infty$), enlarging $F_k$. Let $n_k = n_{k-1} + 1$. Proceed to stage $k+1$.

\item \textbf{Case B:} $W_{h_{n_{k-1}}} \neq F_k$. Then the current hypothesis is already wrong for the finite language $F_k$. We may define $L = F_k$: since $F_k$ is finite and $\mathcal{L}$ contains all finite languages, $F_k \in \mathcal{L}$. We extend the text by repeating elements of $F_k$ forever. The learner $M$ has output a hypothesis $h_{n_{k-1}}$ with $W_{h_{n_{k-1}}} \neq F_k = L$, and the text for $L$ can be arranged so that $M$ never converges to a correct index (we continue to the next stage rather than stopping, as shown below).
\end{itemize}

The key observation is the \emph{diagonalization}: we never let the learner rest.

If Case~A occurs at every stage, then every element of $L_\infty$ is eventually presented, so $t$ is a text for $L_\infty$. At each stage, $M$'s current hypothesis $h_{n_k}$ is checked: does $W_{h_{n_k}} = F_k$? If yes, we expand. But $F_k \subsetneq L_\infty$ for all $k$ (since $L_\infty$ is infinite), so $W_{h_{n_k}} = F_k \neq L_\infty$. Therefore $M$ outputs a hypothesis that is wrong at every stage, it cannot converge to a correct index for $L_\infty$.

If Case~B occurs at some stage $k$, then $M$ has already failed for $F_k$. We commit to $L = F_k$ and repeat elements of $F_k$ forever. But we can do more: before committing, we wait to see if $M$ changes its mind. If $M$ later outputs a hypothesis $h'$ with $W_{h'} = F_k$, we switch to Case~A and add the next element of $L_\infty$. This forces $M$ to fail for $L_\infty$ instead.

In either case, $M$ fails. Since $M$ was an arbitrary learner, no learner $\Ex$-identifies $\mathcal{L}$.
\end{proof}

The proof has a recursive structure that repays careful study. The adversary maintains a ``threat'' at every stage: either the current language is the finite set seen so far, or it is the infinite language $L_\infty$. Whenever the learner commits to one possibility, the adversary switches to the other. The learner is forced to change its mind infinitely often, and $\Ex$-identification requires that it eventually stop changing its mind.

\begin{remark}[Diagonalization vs.\ concentration]
\label{rmk:diag-vs-conc}
Compare this proof to the lower bound for PAC learning (\Cref{ch:pac}). The PAC lower bound constructs a \emph{distribution} that fools the learner with positive probability; it is probabilistic. Gold's proof constructs a \emph{text} that fools the learner with certainty; it is adversarial. The PAC proof uses counting (how many hypotheses can be distinguished by $m$ samples); Gold's proof uses the halting problem implicitly (the adversary must check whether $W_e = F$ for arbitrary $e$). These are entirely different mathematical worlds.
\end{remark}

\begin{historical}
The proof technique of Theorem~\ref{thm:gold} is a \emph{priority argument} in the sense of recursion theory, though a simple one. More sophisticated priority arguments appear in the study of learning with resource bounds (Case and Smith~\cite{CaseSmith1983}). The connection between Gold-style learning and recursion theory runs deep: Barzdin\v{s}~\cite{Barzdinsh1974} showed that the class of languages $\Ex$-identifiable from informant is exactly the class of $\Sigma^0_2$-definable families, establishing a precise link to the arithmetical hierarchy.
\end{historical}

% ================================================================
% SECTION 3: The Hierarchy: BC > Ex > FIN
% ================================================================

\section{The Identification Hierarchy}
\label{sec:hierarchy}

Gold's impossibility theorem says that $\Ex$-identification is limited. Two natural responses: strengthen the success criterion (demand faster convergence) or weaken it (demand less of the final hypothesis). Both directions are fruitful.

\subsection{Finite Identification}
\label{sec:fin-learning}

\begin{defn}[Finite Identification ($\FIN$)]
\label{def:fin-learning}
A learner $M$ \emph{$\FIN$-identifies} a language $L$ from text if, for every text $t$ for $L$, there exists a time $n_0$ such that $M$ outputs exactly one hypothesis, $M$ outputs ``?'' (no guess) for $n < n_0$ and outputs a single index $e$ at time $n_0$ with $W_e = L$, never changing its mind. A class $\mathcal{L}$ is \emph{$\FIN$-identifiable} if a single learner $\FIN$-identifies every $L \in \mathcal{L}$.
\end{defn}

$\FIN$ is the strongest identification criterion: the learner must produce the correct answer in finitely many steps with no subsequent revision. The class of $\FIN$-identifiable families is a strict subset of $\Ex$-identifiable families; for instance, the class of all finite languages is $\Ex$-identifiable (\Cref{ex:finite-languages}) but not $\FIN$-identifiable, since the learner cannot know when the last element has been seen.

\subsection{Behaviorally Correct Learning}
\label{sec:bc-learning}

\begin{defn}[Behaviorally Correct Learning ($\BC$)]
\label{def:bc-learning}
A learner $M$ \emph{$\BC$-identifies} a language $L$ from text if, for every text $t$ for $L$, there exists a time $n_0$ such that $W_{M(t_0, \ldots, t_n)} = L$ for all $n \geq n_0$. That is, from time $n_0$ onward, every hypothesis the learner outputs is \emph{extensionally correct} (computes the right language), but the indices may keep changing.
\end{defn}

The distinction between $\Ex$ and $\BC$ is subtle but consequential. $\Ex$ requires \emph{syntactic} convergence: the index stabilizes. $\BC$ requires only \emph{semantic} convergence: the computed language stabilizes, even if the learner keeps switching between different programs that all compute the same language. At first glance, this seems like a minor relaxation. It is not.

\begin{thm}[Case--Smith {\cite{CaseSmith1983}}]
\label{thm:bc-strictly-stronger}
$\BC$ is strictly more powerful than $\Ex$: there exists a class of languages that is $\BC$-identifiable but not $\Ex$-identifiable from text.
\end{thm}

\begin{proof}
We construct a class $\mathcal{L}$ that separates $\BC$ from $\Ex$.

For each total recursive function $f : \N \to \N$, define the language $L_f = \{\langle n, f(n)\rangle : n \in \N\}$, where $\langle \cdot, \cdot \rangle$ is a computable pairing function. Let $\mathcal{L} = \{L_f : f \text{ is total recursive}\}$.

\medskip\noindent\textbf{$\mathcal{L}$ is $\BC$-identifiable.} Given a text $t$ for some $L_f$, at time $n$ the learner has seen finitely many pairs $\langle n_i, m_i\rangle$. Define the partial function $g_n$ by $g_n(n_i) = m_i$ for all pairs seen so far, and $g_n(k) = 0$ for all other $k$. Output an index for the language $L_{g_n} = \{\langle k, g_n(k)\rangle : k \in \N\}$. Once all pairs $\langle k, f(k)\rangle$ for $k \leq K$ have appeared (for any $K$), the hypothesis computes $L_f$ correctly on $\{0, \ldots, K\}$. As $n \to \infty$, $g_n$ converges pointwise to $f$. Since $f$ is total, for every $k$ there exists a time after which $g_n(k) = f(k)$. The language computed by the hypothesis is eventually $L_f$, though the \emph{index} keeps changing as new pairs are absorbed. This is $\BC$-identification.

\medskip\noindent\textbf{$\mathcal{L}$ is not $\Ex$-identifiable.} Suppose for contradiction that a learner $M$ $\Ex$-identifies $\mathcal{L}$. We diagonalize against $M$. Define a total recursive function $f$ as follows.

Begin presenting pairs $\langle 0, 0\rangle, \langle 1, 0\rangle, \langle 2, 0\rangle, \ldots$ (corresponding to the zero function). Wait until $M$ converges to some index $e$. Since $M$ $\Ex$-identifies $\mathcal{L}$ and the zero function is total recursive, $M$ must converge. Now $W_e = L_{\mathbf{0}}$ (the language of the zero function).

Modify $f$: set $f(k) = 1$ for some large $k$ not yet presented. This changes the target to a different total recursive function, but the text presented so far is consistent with both targets. The learner $M$ has already committed to index $e$, which computes $L_{\mathbf{0}} \neq L_f$. By a careful effectivization of this argument (choosing $k$ computably based on $M$'s behavior), we construct $f \in \mathcal{L}$ on which $M$ fails. This contradicts $M$ $\Ex$-identifying all of $\mathcal{L}$.

The essential point is that $\BC$-identification does not require the \emph{index} to stabilize, only the \emph{extension}. The class $\mathcal{L}$ exploits this: the learner must continually update its program as new pairs arrive, and the programs keep changing, but they all eventually compute the same language. $\Ex$-identification cannot tolerate this perpetual updating of indices.
\end{proof}

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    criterion/.style={draw, thick, rounded corners, minimum width=2.4cm, minimum height=0.9cm, font=\small\bfseries, align=center},
    branch/.style={draw, rounded corners, minimum width=2cm, minimum height=0.7cm, font=\small, align=center},
    strict/.style={-{Stealth[length=2mm]}, thick, sepgreen!70!black},
    weak/.style={-{Stealth[length=2mm]}, thin, dashed, notegray},
    witlabel/.style={font=\tiny, text=edgeorange, fill=white, inner sep=1.5pt, align=center},
    node distance=1.2cm
]

% Main chain
\node[criterion, fill=defblue!15] (bc) {$\BC$};
\node[criterion, fill=defblue!15, below=2cm of bc] (ex) {$\Ex$};
\node[criterion, fill=defblue!15, below=2cm of ex] (fin) {$\FIN$};

% Strict inclusions
\draw[strict] (fin) -- node[witlabel, right, xshift=3pt] {all finite langs\\are $\Ex$ not $\FIN$} (ex);
\draw[strict] (ex) -- node[witlabel, right, xshift=3pt] {Case--Smith 1983\\(Thm~\ref{thm:bc-strictly-stronger})} (bc);

% Branches off Ex
\node[branch, fill=edgeorange!8, right=2.5cm of ex] (anom) {Anomalous\\$\Ex^*$};
\node[branch, fill=edgeorange!8, left=2.5cm of ex] (mono) {Monotonic\\$\mathrm{Mon}\Ex$};
\node[branch, fill=goldcolor!10, below right=1cm and 2.5cm of ex] (vac) {Vacillatory\\$\mathrm{Vex}$};
\node[branch, fill=goldcolor!10, above right=0.5cm and 2.5cm of bc] (trial) {Trial \&\\Error};

% Relations
\draw[strict] (ex) -- node[witlabel, above] {\rel{restricts}} (anom);
\draw[weak] (mono) -- node[witlabel, above] {\rel{restricts}} (ex);
\draw[strict] (ex) -- node[witlabel, below right, xshift=-8pt] {\rel{restricts}} (vac);
\draw[weak] (bc) -- node[witlabel, above right, xshift=-5pt] {$\Delta^0_2$} (trial);

\end{tikzpicture}
\caption{The identification hierarchy. Solid arrows indicate strict inclusion of identifiable classes (more hypotheses identified at the target). $\FIN \subsetneq \Ex \subsetneq \BC$, with witnesses on each edge. Anomalous learning ($\Ex^*$) extends $\Ex$ by removing the zero-error constraint. Monotonic learning restricts $\Ex$ by forbidding hypothesis retraction. Vacillatory learning sits between $\Ex$ and $\BC$.}
\label{fig:identification-hierarchy}
\end{figure}

\begin{traversal}
\textbf{Path:} $\nde{fin\_learning} \xrightarrow{\rel{strictly\_stronger}} \nde{ex\_learning} \xrightarrow{\rel{strictly\_stronger}} \nde{bc\_learning}$.\\[3pt]
Each arrow is witnessed by an explicit separation. The graph encodes these as \rel{strictly\_stronger} edges with the separation witness as metadata. The hierarchy is not a sequence of increasingly liberal definitions, it is a chain of \emph{theorems}, each proved by constructing a class that one criterion can learn and the other cannot.
\end{traversal}

% ================================================================
% SECTION 4: Relaxations and Their Lattice
% ================================================================

\section{Relaxations of Identification}
\label{sec:relaxations}

The $\FIN$--$\Ex$--$\BC$ chain is the spine of the hierarchy. Several natural relaxations branch off from $\Ex$, each modifying a different aspect of the convergence requirement.

\subsection{Anomalous Learning}
\label{sec:anomalous}

\begin{defn}[Anomalous Learning ($\Ex^*$)]
\label{def:anomalous}
A learner \emph{$\Ex^*$-identifies} a language $L$ if it $\Ex$-converges to an index $e$ such that $W_e$ differs from $L$ on at most finitely many strings. That is, the final hypothesis may contain finitely many errors, finitely many strings incorrectly included or excluded.
\end{defn}

Anomalous learning relaxes $\Ex$ by removing the requirement of exact correctness, replacing it with correctness up to a finite set. In graph terms, this is a \rel{restricts} edge from $\nde{anomalous\_learning}$ to $\nde{ex\_learning}$: the constraint (zero anomalies) is removed, and no new grammatical structure is introduced. The class of $\Ex^*$-identifiable families strictly contains the class of $\Ex$-identifiable families~\cite{CaseSmith1983}.

\subsection{Monotonic Learning}
\label{sec:monotonic}

\begin{defn}[Monotonic Learning ($\mathrm{Mon}\Ex$)]
\label{def:monotonic}
A learner \emph{monotonically} $\Ex$-identifies $L$ if it $\Ex$-identifies $L$ and, whenever it changes its hypothesis from $e$ to $e'$, the new hypothesis is at least as inclusive on the data seen so far: $W_e \cap \{t_0, \ldots, t_n\} \subseteq W_{e'} \cap \{t_0, \ldots, t_n\}$. Once a datum is correctly classified, that classification is never retracted.
\end{defn}

Monotonic learning \emph{strengthens} $\Ex$ by adding a constraint. The class of $\mathrm{Mon}\Ex$-identifiable families is a strict subset of $\Ex$-identifiable families. In the graph, $\nde{monotonic\_learning} \xrightarrow{\rel{restricts}} \nde{ex\_learning}$.

\subsection{Vacillatory Learning}
\label{sec:vacillatory}

\begin{defn}[Vacillatory Learning ($\mathrm{Vex}$)]
\label{def:vacillatory}
A learner \emph{vacillatorily} identifies $L$ if it eventually oscillates among finitely many indices $e_1, \ldots, e_k$, all satisfying $W_{e_i} = L$. The learner never settles on a single index, but all its eventual outputs are extensionally correct and drawn from a finite set.
\end{defn}

Vacillatory learning sits between $\Ex$ (which requires convergence to a single index) and $\BC$ (which allows infinitely many correct indices). It is strictly more powerful than $\Ex$ and strictly less powerful than $\BC$.

\subsection{Trial and Error}
\label{sec:trial-error}

\begin{defn}[Trial-and-Error Learning]
\label{def:trial-error}
A \emph{trial-and-error} predicate for a set $A \subseteq \N$ is a computable function $f : \N \to \{0, 1\}$ such that $\lim_{s \to \infty} f_s(n)$ exists for all $n$ and equals the characteristic function of $A$. The learner may ``retract'' previous outputs: it learns by making mistakes and correcting them.
\end{defn}

Trial-and-error learning connects identification in the limit to the arithmetical hierarchy. A set $A$ is trial-and-error learnable if and only if $A \in \Delta^0_2$, the class of sets that are both $\Sigma^0_2$ and $\Pi^0_2$. This is the precise recursion-theoretic characterization, due to Putnam~\cite{Gold1965} and Gold: the limit-computable functions are exactly the $\Delta^0_2$ functions. This connection places Gold-style learning firmly within the landscape of classical recursion theory.

% ================================================================
% SECTION 5: Mind-Change Complexity and Ordinals
% ================================================================

\section{Mind-Change Complexity}
\label{sec:mind-change}

Gold's impossibility theorem tells us that certain classes cannot be identified at all. For classes that \emph{can} be identified, a natural quantitative question arises: how many times must the learner change its mind before converging?

\begin{defn}[Mind Change]
\label{def:mind-change}
  \lean{EXLearnable}
  \leanok
A \emph{mind change} occurs at time $n$ if the learner's output satisfies $M(t_0, \ldots, t_n) \neq M(t_0, \ldots, t_{n-1})$. The \emph{mind-change count} of $M$ on text $t$ is the number of times $M$ changes its hypothesis.
\end{defn}

For finite mind-change bounds, the theory is straightforward: we say that $M$ identifies $\mathcal{L}$ with at most $k$ mind changes if, on every text for every $L \in \mathcal{L}$, the mind-change count is at most $k$. The class of all finite languages is identifiable with 0 mind changes after the first hypothesis (the learner in \Cref{ex:finite-languages} changes its mind every time a new element appears, but a more careful learner can be designed with bounded mind changes for restricted subclasses).

The surprise, genuinely unexpected for readers coming from PAC theory, is that integer-valued bounds are \emph{not sufficient} to characterize the full landscape.

\begin{thm}[Freivalds--Smith {\cite{FreivaldsSmith1993}}]
\label{thm:mind-change-ordinal}
  \lean{mind_change_characterization}
  \leanok
  \uses{def:ex-learning,def:mind-change}
The mind-change complexity of $\Ex$-identification is naturally measured by \emph{countable ordinals}. Specifically:
\begin{enumerate}[label=(\roman*),leftmargin=*]
\item For every countable ordinal $\alpha$, one can define what it means for a learner to identify a class with mind-change bound $\alpha$.
\item There exist classes identifiable with mind-change bound $\omega$ (the first infinite ordinal) that cannot be identified with any finite mind-change bound.
\item More generally, for every ordinal $\alpha < \omega_1$, there exist classes identifiable with mind-change bound $\alpha$ but not with any bound $\beta < \alpha$.
\end{enumerate}
\end{thm}

The idea behind ordinal mind-change bounds is as follows. A learner with mind-change bound $\omega$ begins with an ordinal counter set to $\omega$. At each mind change, the counter must decrease, but it may decrease to any smaller ordinal. Since there is no infinite descending sequence of ordinals (ordinals are well-ordered), the learner must eventually stop changing its mind. The counter $\omega$ allows finitely many mind changes, but the \emph{number} of mind changes need not be bounded in advance by any fixed integer, it may depend on the input.

This is fundamentally different from an integer bound. With a bound of $k \in \N$, the learner may change its mind at most $k$ times on \emph{every} input. With a bound of $\omega$, the learner may change its mind $k$ times for input-dependent $k$, with no uniform finite upper bound. The ordinal hierarchy continues: $\omega \cdot 2$ allows the learner to ``reset'' its finite counter once; $\omega^2$ allows nested levels of resetting; and so on up through the constructive ordinals.

\begin{remark}[Ordinals in learning theory]
\label{rmk:ordinals-surprise}
The appearance of transfinite ordinals in learning theory is a genuine structural surprise. PAC learning theory uses real-valued parameters ($\varepsilon, \delta, m(\varepsilon, \delta)$). Online learning uses integer-valued dimensions ($\Ldim$). Gold-style learning requires ordinal-valued complexity measures. Each proof technique brings its own number system. The full development of ordinal mind-change complexity is deferred to \Cref{ch:ordinals}, where it interacts with the constructive ordinal notation systems of Kleene.
\end{remark}

\begin{traversal}
\textbf{Path:} $\nde{mind\_change\_characterization} \xrightarrow{\rel{measures}} \nde{ex\_learning}$.\\[3pt]
The mind-change ordinal is a complexity measure on $\Ex$-identifiable classes, analogous to the VC dimension for PAC-learnable classes. But where VC dimension is a single integer, mind-change complexity is an ordinal, potentially transfinite. This is a \rel{measures} edge of a qualitatively different kind than $\nde{vc\_dimension} \xrightarrow{\rel{measures}} \nde{concept\_class}$.
\end{traversal}

% ================================================================
% SECTION 6: The Three-Paradigm Separation
% ================================================================

\section{Three Paradigms, Incomparable}
\label{sec:three-paradigm}

We now have three learning paradigms: PAC (\Cref{ch:pac}), online (\Cref{ch:online}), and Gold. Each defines ``learnable'' differently. The question is: what is the relationship?

The answer is that they are \emph{pairwise incomparable}. No paradigm subsumes another. There exist classes learnable under one criterion but not another, in every direction. This is the most important structural fact about the landscape of learning theory.

\begin{separation}
\textbf{Theorem} (Three-Paradigm Separation). \textit{The following four separations hold:}

\begin{enumerate}[label=(\alph*),leftmargin=*,itemsep=4pt]
\item \textbf{$\Ex$-learnable but not PAC-learnable.} The class $\mathcal{L}_{\mathrm{fin}}$ of all finite subsets of $\N$ is $\Ex$-identifiable from text (\Cref{ex:finite-languages}). But the associated concept class has infinite VC dimension, any finite set of points can be shattered, so it is not PAC-learnable.

The witness exploits the fundamental difference in how the two paradigms treat ``all finite subsets.'' $\Ex$-identification succeeds because on any \emph{fixed} finite set, the learner eventually sees all elements. PAC fails because the learner must handle \emph{all} finite sets simultaneously with bounded sample complexity, and the VC dimension is infinite.

\item \textbf{PAC-learnable but not $\Ex$-identifiable.} The class of threshold functions $\mathcal{C} = \{x \mapsto \ind[x \geq \theta] : \theta \in \R\}$ over $\R$ is PAC-learnable ($\VC = 1$). But it is not $\Ex$-identifiable from text, because a text for a threshold language $\{\theta, \theta+1, \theta+2, \ldots\}$ reveals the threshold only in the limit, and the class of all such languages together with all finite languages triggers Gold's impossibility theorem.

\item \textbf{Online-learnable but not $\Ex$-identifiable.} The class of singletons $\mathcal{C} = \{\{x\} : x \in X\}$ has Littlestone dimension~1 (online-learnable with at most 1 mistake). But the class of all singletons together with the empty set is not $\Ex$-identifiable from text for reasons analogous to Gold's theorem: the learner cannot distinguish ``no more positive examples will come'' from ``the next positive example has not yet arrived.''

\item \textbf{$\Ex$-identifiable but not online-learnable.} Consider any class $\mathcal{L}$ of recursive languages that is $\Ex$-identifiable and has infinite Littlestone dimension. Such classes exist: the class of all pattern languages (Angluin, 1980) is $\Ex$-identifiable from text but has infinite Littlestone dimension over suitable domains.
\end{enumerate}
\end{separation}

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    territory/.style={draw, thick, rounded corners=8pt, minimum width=3.5cm, minimum height=2.5cm, font=\small\bfseries, align=center},
    witness/.style={font=\scriptsize, text=edgeorange, fill=white, inner sep=2pt, align=center},
    arrow/.style={-{Stealth[length=2mm]}, thick, thmred, dashed}
]

% Three paradigms as territories
\node[territory, fill=defblue!8] (pac) at (0,0) {PAC\\{\scriptsize $\varepsilon$-$\delta$, distributions}\\{\scriptsize VC dimension}};
\node[territory, fill=goldcolor!8] (gold) at (5,0) {Gold\\{\scriptsize convergence, no $\varepsilon$}\\{\scriptsize ordinal complexity}};
\node[territory, fill=onlineblue!8] (online) at (2.5,3.5) {Online\\{\scriptsize adversarial, regret}\\{\scriptsize Littlestone dim}};

% Separation arrows (does_not_imply)
\draw[arrow] (pac) -- node[witness, below] {thresholds:\\PAC yes, $\Ex$ no} (gold);
\draw[arrow] (gold) -- node[witness, above] {all finite sets:\\$\Ex$ yes, PAC no} (pac);

\draw[arrow] (online) -- node[witness, left, xshift=-2pt] {singletons+$\emptyset$:\\Online yes, $\Ex$ no} (gold);
\draw[arrow] (gold) -- node[witness, right, xshift=2pt] {patterns:\\$\Ex$ yes, Online no} (online);

\draw[arrow] (pac) -- node[witness, left, xshift=-3pt] {thresholds:\\PAC yes,\\Online no ($\Ldim=\infty$)} (online);
\draw[arrow] (online) -- node[witness, right, xshift=3pt] {countable $\Ldim<\infty$,\\$\VC=\infty$} (pac);

\end{tikzpicture}
\caption{The three-paradigm separation. Each pair of paradigms is separated in both directions by explicit witnesses. No paradigm subsumes another. The dashed arrows represent \rel{does\_not\_imply} edges in the graph, and each arrow is labelled with its witness class.}
\label{fig:three-paradigm}
\end{figure}

\begin{remark}[What the separations mean]
\label{rmk:separation-meaning}
The three-paradigm separation is not a deficiency of the definitions. It reflects a genuine mathematical fact: different notions of ``learning from data'' capture different aspects of the learning problem, and no single notion is universal. PAC learning measures statistical generalization under distributional assumptions. Online learning measures worst-case sequential prediction. Gold-style identification measures eventual convergence without any accuracy or efficiency guarantee. These are three different mathematical questions, and they have three different answers.
\end{remark}

% ================================================================
% SECTION 7: Chapter Summary
% ================================================================

\section{What This Chapter Established}
\label{sec:ch7-summary}

This chapter introduced Gold-style identification in the limit, the oldest paradigm in formal learning theory, and established five structural facts:

\begin{enumerate}[leftmargin=*,itemsep=3pt]
\item \textbf{Gold's impossibility theorem} (\Cref{thm:gold}): any class containing all finite languages and at least one infinite language is not $\Ex$-identifiable from text. The proof is by diagonalization, the first impossibility result in learning theory, and one whose proof technique (adversarial stream construction) has no analogue in PAC or online theory.

\item \textbf{The identification hierarchy} (\Cref{sec:hierarchy}): the chain $\FIN \subsetneq \Ex \subsetneq \BC$ is strict, with the Case--Smith separation (\Cref{thm:bc-strictly-stronger}) providing the witness for $\BC > \Ex$. The hierarchy is not a sequence of definitions but a chain of theorems.

\item \textbf{Relaxations form a lattice} (\Cref{sec:relaxations}): anomalous, monotonic, and vacillatory learning branch off from $\Ex$, each modifying a different aspect of the convergence requirement. Trial-and-error learning connects to $\Delta^0_2$ in the arithmetical hierarchy.

\item \textbf{Mind-change complexity is ordinal-valued} (\Cref{sec:mind-change}): the Freivalds--Smith characterization shows that the right complexity measure for $\Ex$-identification uses transfinite ordinals, not integers. The full treatment is in \Cref{ch:ordinals}.

\item \textbf{The three paradigms are pairwise incomparable} (\Cref{sec:three-paradigm}): PAC, online, and Gold-style identification are separated in every direction by explicit witnesses. This is the most important structural fact about the landscape of learning theory.
\end{enumerate}

The recursion-theoretic character of this chapter, diagonalization proofs, connections to the arithmetical hierarchy, ordinal-valued measures, contrasts sharply with the probabilistic character of \Cref{ch:pac} and the combinatorial character of \Cref{ch:online}. Each paradigm brings its own mathematical world. The separations of \Cref{sec:three-paradigm} are consequences of this fact: the paradigms use different proof techniques because they formalize genuinely different questions.


% ================================================================
% EXERCISES
% ================================================================

\subsection*{Exercises}

\begin{enumerate}[leftmargin=*,itemsep=6pt]

\item \textbf{The power of informant over text.}
  Gold showed that the class $\mathcal{L}$ of all recursive languages
  is $\Ex$-identifiable from \emph{informant} (both positive and
  negative examples) but not from text (positive examples only).
  \begin{enumerate}[label=(\alph*)]
  \item Prove the informant direction: construct an $\Ex$-learner $M$
    that identifies any recursive language $L$ from informant.
    \emph{Hint:}  Dovetail over all programs $\varphi_0, \varphi_1,
    \ldots$, and at time $t$ hypothesize the smallest index~$e$
    consistent with all labeled data seen so far.  Use the fact that
    the informant eventually presents every string with its correct
    label, so incorrect indices are eventually refuted.
  \item Explain precisely why this learner fails from text.  Identify
    the step in the argument that relies on negative examples, and
    show that no text-based substitute exists.  (\emph{Hint:} The
    learner can refute ``$\varphi_e$ includes $x$'' from an
    informant presentation of $(x, 0)$, but a text for $L$ never
    explicitly says ``$x \notin L$'', it merely fails to present
    $x$, which is indistinguishable from delay.)
  \end{enumerate}

\item \textbf{$\Ex$-identification of co-finite languages.}
  Let $\mathcal{L}_{\mathrm{cof}}$ be the class of all co-finite
  languages over $\N$: $L \in \mathcal{L}_{\mathrm{cof}}$ iff
  $\N \setminus L$ is finite.
  \begin{enumerate}[label=(\alph*)]
  \item Prove that $\mathcal{L}_{\mathrm{cof}}$ is
    $\Ex$-identifiable from text.  (\emph{Hint:} Every co-finite
    language $L$ contains all but finitely many natural numbers.
    A text for $L$ eventually presents every element of $L$, so after
    time $n_0$, every natural number $\leq n_0$ has either appeared
    in the text (and is in $L$) or has not appeared (and may or
    may not be in $L$).  Design a learner that waits long enough
    to conclude that unseen small numbers are \emph{not} in $L$.)
  \item Prove that $\mathcal{L}_{\mathrm{fin}} \cup
    \mathcal{L}_{\mathrm{cof}}$ (all finite and all co-finite
    languages together) is \emph{not} $\Ex$-identifiable from text.
    (\emph{Hint:} Apply Gold's impossibility theorem.  Verify
    that this class contains all finite languages and at least one
    infinite language.)
  \item The result of~(b) is sharper than Gold's theorem applied
    to ``all finite + one infinite'': the single infinite language
    is itself very structured (co-finite).  Explain why the structure
    of the infinite language does not help, what makes the
    diagonalization work is not the complexity of $L_\infty$ but
    the learner's inability to distinguish ``finite set, done
    growing'' from ``co-finite set, still growing.''
  \end{enumerate}

\end{enumerate}

